---
title: "mslp"
output:
  rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{mslp}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo = FALSE, message = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

Mslp is a comprehensive pipeline to predict potential synthetic lethal partners (SLPs) for human tumour mutations. The underlying assumpations are 1) SLPs are up-regulated to compensate the loss of function of mutations, and 2) SLPs correlates with mutations in the wide type patients.

Primary SLPs are first predicted by the compensationModule and correlationModule. Together with the genetic screen data, consensus SLPs are identified as recurrent primary SLPs of same muations in different cell lines. Following parts of the tutorial show a complete run from data preparation to consensus SLPs.

## Data preprocessing
In the tutorial we could use files from PanCancer Atlas, provided by [cBioPortal](https://www.cbioportal.org/) [1].

- mutation profiles, "data_mutations_extended.txt".
- CNA profiles, "data_CNA.txt".
- gene exprssion, "data_RNA_Seq_v2_expression_median.txt".
- z-score data, "data_RNA_Seq_v2_mRNA_median_Zscores.txt".

It is possible to use customized datasets, simply following the format and column names of above files, e.g., "data_mutations_extended.txt" is a gene by sample matrix, with "Hugo_Symbol", "Entrez_Gene_Id" as the first two columns. Three columns are mandatory in mutation profiles: "Tumor_Sample_Barcode", "Gene" and "Variant_Classification", while "Gene" contains the Ensembl gene ids.

```{r}
#- Preprocessing the data.
library(mslp)
library(magrittr)

#- Path to input files.
P_mut  <- "data_mutations_extended.txt"
P_cna  <- "data_CNA.txt"
P_expr <- "data_RNA_Seq_v2_expression_median.txt"
P_z    <- "data_RNA_Seq_v2_mRNA_median_Zscores.txt"

res <- pp_tcga(P_mut, P_cna, P_expr, P_z)

saveRDS(res$mut_data, "mut_data.rds")
saveRDS(res$expr_data, "expr_data.rds")
saveRDS(res$zscore_data, "zscore_data.rds")
```

Here we load toy datasets.

```{r}
#- mutation from TCGA datasets to computate SLP via comp_slp
data(comp_mut)
#- mutation from TCGA datasets to computate SLP via corr_slp. 
data(corr_mut)
data(example_expr)
data(example_z)
```

## Call SLPs from compensationModule
We use `comp_slp` to predict SLPs compensated for  mutation-induced loss of functions. Briefly, it identifies over-expressed genes in patients with interseted mutations via the rank products algorithm [2], while co-occurring mutations are removed beforehand.

```{r}
res <- comp_slp(example_z, comp_mut)
saveRDS(res, file = "compSLP_res.rds")
```

## Call SLPs from correlationModule
We use `corr_slp` to predict SLPs correlated with mutations in wide type patients. Internally GENIE3 is used to select potential SLPs [3]. The im_thresh of 0.0016 was a rather robust value identified from various TCGA datasets with 100 random selected mutations and 50 repetitions.

```{r}
res <- corr_slp(example_expr, corr_mut)
saveRDS(res, "corrSLP_res.rds")

#- Filter res by importance threshold to reduce false positive.
im_thresh    <- 0.0016
res_filtered <- res[im >= im_thresh]
saveRDS(res_filtered, "corrSLP_filtered.rds")
```

It is recommended to compute the `im_thresh` for each cancer type separately. We derived an approach estimating the threshold by running `corr_slp` for randomly selected mutations repeatedly. SLPs of high relevance are picked as "true" SLPs for each mutation using the rank products algorithm. We then calculated the best threshold of receiver operating characteristic curve  (ROC) of each repetation, and take the mean value across repetition. The average value among mutations is the final threshold. 

```{r}
#- Random mutations and runs
mutgene      <- sample(intersect(corr_mut$mut_entrez, rownames(example_expr)), 5)
nperm        <- 3

res          <- lapply(seq_len(nperm), function(x) corr_slp(example_expr, corr_mut, mutgene = mutgene))
im_res       <- est_im(res)
res_filtered <- res[im >= mean(im_res$roc_thresh)]
saveRDS(res_filtered, "corrSLP_filtered.rds")
```

## Call consensus SLPs
The mutation profiles and genetic screen data of cancer cell lines are required for this step. Cancer Cell Line Encyclopedia (CCLE) is a great place to find mutation data [4]; and genetic screen results could be found in datatsets such as Project Drive [5] and DepMap [6].

For example, following codes show how to extract mutation data from CCLE.
```{r}
library(readxl)

#- nature11003-s3.xls is available in the supplmentary data of CCLE publication.
ccle <- readxl::read_excel("nature11003-s3.xls", sheet = "Table S1", skip = 2) %>%
  as.data.frame %>%
  set_names(gsub(" ", "_", names(.))) %>%
  as.data.table %>%
  .[, CCLE_name := toupper(CCLE_name)] %>%
  unique

#- ONly use the Nonsynonymous Mutations. CCLE_DepMap_18Q1_maf_20180207.txt could be downloaded in the CCLE website.
#- Only need the columns of cell_line and mut_entrez.
mut_type  <- c("Missense_Mutation", "Nonsense_Mutation", "Frame_Shift_Del", "Frame_Shift_Ins", "In_Frame_Del", "In_Frame_Ins", "Nonstop_Mutation")
ccle_mut  <- fread("CCLE_DepMap_18Q1_maf_20180207.txt") %>%
  .[Variant_Classification %in% mut_type] %>%
  .[, Tumor_Sample_Barcode := toupper(Tumor_Sample_Barcode)] %>%
  .[, Entrez_Gene_Id := as.character(Entrez_Gene_Id)] %>%
  .[, .(Tumor_Sample_Barcode, Entrez_Gene_Id)] %>%
  unique %>%
  setnames(c("cell_line", "mut_entrez"))

#- Select brca cell lines
brca_ccle_mut <- ccle_mut[cell_line %in% unique(ccle[CCLE_tumor_type == "breast"])]
```

Now we are ready to uncover consensus SLPs, which are 1) hits in genetic screens, 2) consistent for same mutations among cell lines. `scr_slp` and `cons_slp` are used for these two steps, respectively.

```{r}
#- Merge data.
#- Load previous results.
comp_res   <- readRDS("compSLP_res.rds")
corr_res   <- readRDS("corrSLP_res.rds")
merged_res <- merge_slp(comp_res, corr_res)

#- Toy hits data. 
screen_1 <- merged_res[, .(slp_entrez, slp_symbol)] %>%
    unique %>%
    .[sample(nrow(.), round(.8 * nrow(.)))] %>%
    setnames(c(1, 2), c("screen_entrez", "screen_symbol")) %>%
    .[, cell_line := "cell_1"]

screen_2 <- merged_res[, .(slp_entrez, slp_symbol)] %>%
    unique %>%
    .[sample(nrow(.), round(.8 * nrow(.)))] %>%
    setnames(c(1, 2), c("screen_entrez", "screen_symbol")) %>%
    .[, cell_line := "cell_2"]

screen_hit <- rbind(screen_1, screen_2)

#- Toy mutations data.
mut_1 <- merged_res[, .(mut_entrez)] %>%
    unique %>%
    .[sample(nrow(.), round(.8 * nrow(.)))] %>%
    .[, cell_line := "cell_1"]

mut_2 <- merged_res[, .(mut_entrez)] %>%
    unique %>%
    .[sample(nrow(.), round(.8 * nrow(.)))] %>%
    .[, cell_line := "cell_2"]

mut_info <- rbind(mut_1, mut_2)

#- Hits that are identified as SLPs.
scr_res <- lapply(c("cell_1", "cell_2"), scr_slp, screen_hit, mut_info, merged_res)
scr_res[lengths(scr_res) == 0] <- NULL
scr_res <- rbindlist(scr_res)

k_res   <- cons_slp(scr_res, merged_res)
#- Filter results, e.g., by kappa_value and padj.
k_res_f <- k_res[kappa_value >= 0.6 & padj <= 0.1]

sessionInfo()
```

## Reference
[1]: Cerami, E. et al. The cBio Cancer Genomics Portal: An Open Platform for Exploring Multidimensional Cancer Genomics Data. Cancer Discovery 2, 401–404 (2012).

[2]: Breitling, R., Armengaud, P., Amtmann, A. & Herzyk, P. Rank products: a simple, yet powerful, new method to detect differentially regulated genes in replicated microarray experiments. FEBS Letters 573, 83–92 (2004).

[3]: Huynh-Thu, V. A., Irrthum, A., Wehenkel, L. & Geurts, P. Inferring Regulatory Networks from Expression Data Using Tree-Based Methods. PLoS ONE 5, e12776 (2010).

[4]: Barretina, J. et al. The Cancer Cell Line Encyclopedia enables predictive modelling of anticancer drug sensitivity. Nature 483, 603 (2012).

[5]: McDonald, E. R. et al. Project DRIVE: A Compendium of Cancer Dependencies and Synthetic Lethal Relationships Uncovered by Large-Scale, Deep RNAi Screening. Cell 170, 577-592.e10 (2017).

[6]: Tsherniak, A. et al. Defining a Cancer Dependency Map. Cell 170, 564-576.e16 (2017).
