---
title: "mslp"
output:
  rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{mslp}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo = FALSE, message = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

mslp provides a comprehensive pipelines to predict potential synthetic lethal partners (SLPs) for human tumour mutations.
The underlying assumpations are 1) SLPs are up-regulated to compensate the loss of function of mutations,
and 2) SLPs correlates with mutations in the wide type patients.

Primary SLPs are first predicted by the compensationModule and correlationModule. Together with the genetic screen data,
consensus SLPs are identified as recurrent primary SLPs for the same muations in different cell lines.
Following parts of the tutorial show a complete run from data preparation to consensus SLPs.

## Data preprocessing
In the tutorial we will use files from PanCancer Atlas, provided by [cBioPortal](https://www.cbioportal.org/) [1].

- mutation profiles, "data_mutations_extended.txt".
- CNA profiles, "data_CNA.txt".
- gene exprssion, "data_RNA_Seq_v2_expression_median.txt".
- z-score data, "data_RNA_Seq_v2_mRNA_median_Zscores.txt".


It is possible to use customized datasets, simply following the
format and column names of above files, e.g., "data_mutations_extended.txt" is a gene by sample matrix, with "Hugo_Symbol", "Entrez_Gene_Id" as
the first two columns. Three columns are mandatory in mutation profiles: "Tumor_Sample_Barcode", "Gene" and "Variant_Classification",
while "Gene" contains the Ensembl gene ids.

```{r}
#- Preprocessing the data.
library(mslp)

#- Path to input files.
P_mut  <- "data_mutations_extended.txt"
P_cna  <- "data_CNA.txt"
P_expr <- "data_RNA_Seq_v2_expression_median.txt"
P_z    <- "data_RNA_Seq_v2_mRNA_median_Zscores.txt"

res <- pp_tcga(P_mut, P_cna, P_expr, P_z)

saveRDS(res$mut_data, "mut_data.rds")
saveRDS(res$expr_data, "expr_data.rds")
saveRDS(res$zscore_data, "zscore_data.rds")
```

## Call SLPs from compensationModule
We use `comp_slp` to predict SLPs compensated for the loss of function of mutations. Briefly, it identifies highly expressed genes in patients for mutations via the rank products algorithm [2], while co-occured mutations are removed beforehand.

```{r}
zscore_data <- readRDS("zscore_data.rds")
mutcna_data <- readRDS("mut_data.rds")

res <- comp_slp(zscore_data, mutcna_data)
saveRDS(res, file = "compSLP_res.rds")
```

## Call SLPs from correlationModule
We use `corr_slp` to predict SLPs correlated with mutations in wide type patients. Internally, GENIE3, a random forest based methods, is used to select potential SLPs [3].

```{r}
expr_data <- readRDS("expr_data.rds")
mut_data  <- readRDS("mut_data.rds")

res <- corr_slp(expr_data, mut_data)

#- Filter the res by importance threshold
im_thresh <- 0.0016
res_filtered <- res[im >= im_thresh]
```

It is advised to compute the `im_thresh` for individual studies. We implemented an approach estimating a threshold via  `corr_slp` repetition results.
Briefly, `corr_slp` are performed for randomly selected mutations repeatedly. For each mutation,
SLPs with high importance are selected as "true" SLPs via the rank products algorithm.
We then calculated the best threshold of receiver operating characteristic curve  (ROC) of each repetation,
and take the mean value across repetition. The final threshold is the average value among mutations.

```{r}
#- Random mutations and runs
mutgene      <- sample(intersect(mut_data$mut_entrez, rownames(expr_data)), 100)
nperm        <- 50

res          <- lapply(seq_len(nperm), function(x) corr_slp(expr_data, mut_data, mutgene = mutgene))
im_res       <- est_im(res)
res_filtered <- res[im >= mean(im_res$roc_thresh]
saveRDS(res_filtered, "corrSLP_filtered.rds")
```

## Call consensus SLPs
We need to prepare mutation and genetic screen data of cancer cell lines.
Cancer Cell Line Encyclopedia (CCLE) is a great place to find mutation data [4]; and genetic screen results could
be found in datatsets such as Project Drive [5] and DepMap [6].
Screen data of two breast cancer cell lines from Project Drive is included in the package as an example (a data.table).

For example, following codes show how to extract mutation data from CCLE.
```{r}
library(readxl)
library(magrittr)
library(data.table)

#- nature11003-s3.xls is available in the supplmentary data of CCLE publication.
ccle <- readxl::read_excel("nature11003-s3.xls", sheet = "Table S1", skip = 2) %>%
  as.data.frame %>%
  set_names(gsub(" ", "_", names(.))) %>%
  as.data.table %>%
  .[, CCLE_name := toupper(CCLE_name)] %>%
  unique

#- Keep the Nonsynonymous Mutations, CCLE_DepMap_18Q1_maf_20180207.txt could be downloaded in the CCLE website.
#- Only need the columns of cell_line and mut_entrez.
mut_type  <- c("Missense_Mutation", "Nonsense_Mutation", "Frame_Shift_Del", "Frame_Shift_Ins", "In_Frame_Del", "In_Frame_Ins", "Nonstop_Mutation")
ccle_mut  <- fread("CCLE_DepMap_18Q1_maf_20180207.txt") %>%
  .[Variant_Classification %in% mut_type] %>%
  .[, Tumor_Sample_Barcode := toupper(Tumor_Sample_Barcode)] %>%
  .[, Entrez_Gene_Id := as.character(Entrez_Gene_Id)] %>%
  .[, .(Tumor_Sample_Barcode, Entrez_Gene_Id)] %>%
  unique %>%
  setnames(c("cell_line", "mut_entrez"))

#- Select brca cell lines
brca_ccle_mut <- ccle_mut[cell_line %in% unique(ccle[CCLE_tumor_type == "breast"])]
```

<br />

Now we are ready to find consensus SLPs, which are 1) hits in genetic screens, 2) consistent for same mutations among cell lines.
`scr_slp` and `cons_slp` are used for these two steps, respectively.

```{r}
#- Merge data.
comp_res   <- readRDS("compSLP_res.rds")
corr_res   <- readRDS("corrSLP_filtered.rds")
merged_res <- merge_slp(comp_res, corr_res)

data(brca_screen)
brca_screen
#       screen_entrez screen_symbol     cell_line
#    1:          8086          AAAS  JIMT1_BREAST
#    2:            15         AANAT HS578T_BREAST
#    3:        154664        ABCA13 HS578T_BREAST
#    4:         10347         ABCA7 HS578T_BREAST
#    5:            22         ABCB7 HS578T_BREAST
#   ---
# 1868:          9406        ZRANB2  JIMT1_BREAST
# 1869:         80345       ZSCAN16 HS578T_BREAST
# 1870:         11130         ZWINT HS578T_BREAST
# 1871:         11130         ZWINT  JIMT1_BREAST
# 1872:         26009          ZZZ3  JIMT1_BREAST

#- Hits that are SLPs as well.
allcell <- intersect(brca_ccle_mut$cell_line, brca_screen$cell_line)
scr_res <- lapply(allcell, scr_slp, brca_screen, cell_mut, merged_res)
scr_res[lengths(scr_res) == 0] <- NULL
scr_res <- rbindlist(scr_res)

#- consensus SLPs by kappa test.
k_res <- cons_slp(scr_res, tumour_slp = merged_res)

#- Filter results, e.g., by kappa_value and padj.
k_res_f <- k_res[kappa_value >= 0.6 & padj <= 0.1]
```

## Reference
[1]: Cerami, E. et al. The cBio Cancer Genomics Portal: An Open Platform for Exploring Multidimensional Cancer Genomics Data. Cancer Discovery 2, 401–404 (2012).

[2]: Breitling, R., Armengaud, P., Amtmann, A. & Herzyk, P. Rank products: a simple, yet powerful, new method to detect differentially regulated genes in replicated microarray experiments. FEBS Letters 573, 83–92 (2004).

[3]: Huynh-Thu, V. A., Irrthum, A., Wehenkel, L. & Geurts, P. Inferring Regulatory Networks from Expression Data Using Tree-Based Methods. PLoS ONE 5, e12776 (2010).

[4]: Barretina, J. et al. The Cancer Cell Line Encyclopedia enables predictive modelling of anticancer drug sensitivity. Nature 483, 603 (2012).

[5]: McDonald, E. R. et al. Project DRIVE: A Compendium of Cancer Dependencies and Synthetic Lethal Relationships Uncovered by Large-Scale, Deep RNAi Screening. Cell 170, 577-592.e10 (2017).

[6]: Tsherniak, A. et al. Defining a Cancer Dependency Map. Cell 170, 564-576.e16 (2017).
